---
title: "Populist parties: still a home for the politically distrustful, or a faded and institutionalized phenomenon?"
author: "Giovanni Bacchiega"
date: today
format:
  html:
    theme: default
    fontsize: 12pt
    linestretch: 2
    toc: true
bibliography: references.bib
citeproc: true
csl: "https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl"
---

```{r}
#| include: FALSE
#| label: General library

#Library needed

library(haven) #reading dta file
library(tidyverse)
library(lubridate) #used to extract year from data_nascita
library(stargazer)
library(sjPlot)
```

```{r}
#| include: FALSE
#| label: FULL CODE

#load the ITANES database & keep interest VARs----

ITANES_2022_Post_electoral <- read_dta("Itanes2022_POST_release 01_weighted.dta")

ITANES_2022_Post_electoral <- ITANES_2022_Post_electoral%>%
  mutate(
    data_pulita = dmy(data_nascita),
    anno_nascita = year(data_pulita),
    anno_nascita = if_else(anno_nascita > 2022, anno_nascita - 100, anno_nascita),
    eta = 2022 - anno_nascita #create age variable 
  )

#INDEX CREATION----

#We want to create a index for political trust, we will therefore run a FA, on selected trust
#related items from ITANES data to check wether the items are related

#Load required packages

library(tidyverse)
library(corrplot)
library(psych)
pca_library <- c("corrr", "ggcorrplot", "FactoMineR", "factoextra")

lapply(pca_library, library, character.only = TRUE)

#Keep just relevant VARs

trust_data_NA <- ITANES_2022_Post_electoral%>%
  select(seriale, Q04b, Q05b, Q05c, Q07a, Q07b, Q07e)%>%
  mutate(across(-seriale, ~ ifelse(.x >= 50, NA, .x))) %>%
  mutate( #reverse were needed to put the negative connotation on the high end
    Q04b = 6 - Q04b,
    Q05b = 6 - Q05b,
    Q05c = 6 - Q05c,
  )

# NA imputation

# Here NA imputation is quite important: we can't just remove NA as we probably
# are in a MNAR (missing not at random) situation, moreover these are a lot of 
# items, it would be quite bad to lose all the informations for just 1 NA
# We can use mice (multiple imputation by chained equations) that take the most
# similar observation which have a value in the variable of interest

library(mice)

imputation <- mice(trust_data_NA,
                   m = 1,
                   method = "pmm",
                   seed = 123,
                   printFlag = FALSE)

trust_data <- complete(imputation, 1)


#Transform Q06 as numeric

# Create a var vector for convenience

cat_var <- c("Q04b", "Q05b", "Q05c", "Q07a", "Q07b","Q07e") 

# With mixcor we impute a mixed correlation matrix

mixed_cormat <- mixedCor(data = trust_data,
                         c = cat_var,
                         smooth = TRUE)

# Plot the correlation 

corrplot(mixed_cormat$rho, method = "number",
         type = "lower", order = "hclust",
         tl.col = "black", tl.srt = 45)

# Since there are a lot of strong correlation use pca to determine how many 
# dimension to use 

eigenvalues <- eigen(mixed_cormat$rho)$values

explained_variance <- (eigenvalues / sum(eigenvalues)) * 100

df_screeplot <- data.frame(
  Component = 1:length(eigenvalues),
  Eigenvalues = eigenvalues,
  Variance <- explained_variance
)

print(df_screeplot)

# We have 3 component with >1 eigenvalue, so for the Kaiser method (keeping PCs)
# with EV > 1 we would take 3 of them. Anyway, we can also use a screeplot to visualize it

ggplot(df_screeplot, aes(x = as.factor(Component), y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.6) +
  geom_line(aes(x = Component, y = Variance), color = "black", linewidth = 0.8) +
  geom_point(aes(x = Component, y = Variance), color = "red", size = 3) +
  geom_hline(yintercept = (1/length(eigenvalues))*100, linetype = "dashed", color = "darkgray") +
  geom_text(aes(label = round(Variance, 1)), vjust = -1.5, size = 3.5) +
  labs(title = "Scree Plot",
       subtitle = "Elbow method",
       x = "Components",
       y = "Explained variance") +
  theme_minimal()

# The screeplot shows a sharp cut between first and second component, suggesting that
# We can try both 1 & 3 components

pca_model <- principal(r = mixed_cormat$rho,
                       nfactors = 1,
                       rotate = "varimax")

pca_model$loadings
pca_model$uniquenesses

pca_model2 <- principal(r = mixed_cormat$rho,
                        nfactors = 2,
                        rotate = "varimax")

pca_model2$loadings
pca_model2$uniquenesses
pca_model2$weights


# Perform FA with 2 dimension 
fa <- fa(r = mixed_cormat$rho,
         nfactors = 2,
         rotate = "varimax")

# Summarize fa and extract loadings & scores
summary(fa)
fa$loadings
fa$uniquenesses
fa$weights

# We now discovered 2 indexes: institutional distrust (MR1) & Stealth democracy attitudes (MR2)
# Stealth democracy: want the system to work well but don't want the politics to 
# lead it

# I WON'T INCLUDE STEALTH DEMOCRACY IN MY COMPUTATION

#Create the new weighted indexes & scale it 0-10

trust_data <- trust_data %>%
  mutate(Institutional_distrust = Q04b*-0.002 + Q05b*-0.036 + Q07a*0.501 + 
           Q05c*-0.006 + Q07b*0.331 + Q07e*0.211)

#we determine a function to rescale -5 +5

rescale <- function(x) {
  x_numeric <- as.numeric(x) #force x to be numeric
  
  x_min <- min(x_numeric, na.rm = TRUE)
  x_max <- max(x_numeric, na.rm = TRUE)
  
  (((x_numeric - x_min) * 10) / (x_max - x_min))
}

trust_rescaled <- trust_data %>%
  mutate(Institutional_distrust = rescale(Institutional_distrust),
  )%>%
  select(seriale, Institutional_distrust)


#FINAL DF ----

populist_df <- ITANES_2022_Post_electoral%>%
  select(seriale, eta_classe, Q10LHa, Q10LHb,eta)%>%
  mutate(seriale = as.numeric(zap_labels(seriale)),
         abstention = as.numeric(zap_labels(Q10LHa)))%>%
  filter(abstention < 3)%>%
  mutate(abstention = ifelse(abstention == 2, 1, 0),
         abstention = as.factor(abstention))%>%
  mutate(populist_vote = Q10LHb %in% c("2","5"))%>%
  mutate(eta_classe = fct_collapse( #merging classes 1 & 2
    factor(eta_classe),
    "18-34" = c("1", "2"),  
    "35-44"   = "3",
    "45-54"   = "4",
    "55-64"   = "5",
    ">64"   = "6"
  ))%>%
  mutate(age_class = eta_classe)%>%
  select(seriale, age_class, populist_vote, eta, abstention)


pop_age <- left_join(trust_rescaled, populist_df, by = join_by(seriale))

table(pop_age$age_class)     


#Addittive models-----

pop_age$populist_vote <- as.factor(pop_age$populist_vote)

Complete_pooling <- glm(populist_vote ~ Institutional_distrust,
                        family = binomial(link = "logit"),
                        data = pop_age)

stargazer(Complete_pooling, type = "text")

plot_model(Complete_pooling, 
           type = "pred", 
           terms = c("Institutional_distrust"),
) +
  theme_minimal()

#Try it on vote/non vote

CP_abstention <- glm(abstention ~ Institutional_distrust,
                     family = binomial(link = "logit"),
                     data = pop_age)

stargazer(CP_abstention, type = "text")

plot_model(CP_abstention, 
           type = "pred", 
           terms = "Institutional_distrust",
) +
  theme_minimal()

# GAM - Check the shape of the distribution ----

# A gam is a model that guarantee a high flexibility & calculate non linear regression models
# we want to use it to study the shape of the relation

library(mgcv)

mod_gam <- gam(populist_vote
               ~ s(Institutional_distrust)
               + age_class,
               family = binomial, data = pop_age)

plot(mod_gam, pages = 1, shade = TRUE)

summary(mod_gam)

# It seems like at a certain point the populist vote decrease for institutional distrust
# This can suggest that after a certain ceiling non voting become the choice: this could
# be the reason for our non working models. We can try a model that take institutional
# distrust till 7/8 and check the significativity of the relation and we can do the same
# for non vote. Moreover we can impute a GAM for abstention


mod_gam <- gam(abstention
               ~ s(Institutional_distrust)
               + age_class,
               family = binomial, data = pop_age)

plot(mod_gam, pages = 1, shade = TRUE)

# Our HP was right abstention explode for really high levels of distrust 
# Now we can try our CDT (critical distrust threshold) model with level lower than 8

# Before starting with our plan we want to check how many high distrust individuals we have
table(pop_age$Institutional_distrust >= 8)

# 987 is really significant, this means that they have a strong impact on our regression models

# CDT simple model ----

CDT_df <- pop_age%>%
  filter(Institutional_distrust <= 8)

CDT <- glm(populist_vote ~ Institutional_distrust,
                        family = binomial(link = "logit"),
                        data = CDT_df)

stargazer(CDT, type = "text")

plot_model(CDT, 
           type = "pred", 
           terms = "Institutional_distrust [all]",
           
) +
  theme_minimal()

# INTERACTION MODEL ----

# Clean the dataset to perform interaction model and avoid confusing sjPlot

pop_age_clean <- pop_age %>%
  filter(Institutional_distrust > 0)%>%
  drop_na() %>%
  mutate(
    log_distrust = log1p(Institutional_distrust), #without transforming them before sjPlot wouldn't recognize them so we have to perform this now
    age_class   = as.factor(age_class)
  )

# INSTITUIONAL DISTRUST X AGE CLASS

# I have to decide a reference class. For my purpose the main strategy would be to choose 
# 18-34 to see its behaviour confronted with the others 

table(pop_age_clean$age_class)  

#18-34 35-44 45-54 55-64   >64 
#756   701   909   852    1269 

# A solution could be to merge 1 & 2

pop_age_clean$age_class <- relevel(factor(pop_age_clean$age_class), ref = "18-34")

distrust_x_class <- glm(populist_vote
                        ~ Institutional_distrust
                        * age_class,
                        family = binomial(link = "logit"),
                        data = pop_age_clean)

stargazer(distrust_x_class, type = "text")

sjPlot::plot_model(distrust_x_class, 
                   type = "int",
                   show.legend = TRUE
) +
  theme_bw()

distrust_x_class_abs <- glm(abstention
                            ~ Institutional_distrust
                            * age_class,
                            family = binomial(link = "logit"),
                            data = pop_age_clean)

stargazer(distrust_x_class_abs, type = "text")

sjPlot::plot_model(distrust_x_class_abs, 
                   type = "int",
                   show.legend = TRUE
) +
  theme_bw()


# CDT Model -----

# CDT addictive

CDT <- glm(populist_vote
             ~ Institutional_distrust,
             family = binomial(link = "logit"),
             data = CDT_df)

stargazer(CDT, type = "text")

plot_model(CDT,
           type = "pred",
           terms = "Institutional_distrust",
           show.legend = TRUE) + 
  theme_minimal()

# CDT Interaction 

CDT_df$age_class <- relevel(factor(CDT_df$age_class), ref = "18-34")

CDT_distrust_int <- glm(populist_vote
                          ~ Institutional_distrust
                          * age_class,
                          family = binomial(link = "logit"),
                          data = CDT_df)

stargazer(CDT_distrust_int, type = "text")

plot_model(CDT_distrust_int, 
           type = "int",
           ci.lvl = NA,
           show.legend = TRUE,
) +
  theme_bw()

table(pop_age_clean$age_class)     


# Control by FDI
# Control by macro-regions / instruction 
# Control by M5S/League

#Macro regions controlling ----

# North - South - Red zone (included: Toscana, Emilia-Romagna, Umbria, Marche)

Regions_df <- ITANES_2022_Post_electoral%>%
  mutate(Macro_region = case_when(
    regione < 7 ~ "North",
    regione < 12 ~ "Red_area",
    TRUE ~ "South"))%>%
  select(seriale, Macro_region)

Regions_df <- left_join(pop_age_clean, Regions_df, by = join_by(seriale))%>%
  mutate(Macro_region = factor(Macro_region))

# Selection of the reference class

Regions_df$age_class <- relevel(factor(Regions_df$age_class), ref = "18-34")

#Pure addittive model of populist vote with control

Regions_model_add <- glm(populist_vote
                         ~ Macro_region + Institutional_distrust
                         + age_class,
                         family = binomial(link = "logit"),
                         data = Regions_df)

stargazer(Regions_model_add, type = "text")

plot_model(Regions_model_add, 
           type = "pred", 
           terms = c("Institutional_distrust", "age_class"),
) +
  theme_minimal()

# Pure addittive model of abstention with control


Regions_model_add_abs <- glm(abstention
                             ~ Macro_region + Institutional_distrust
                             + age_class,
                             family = binomial(link = "logit"),
                             data = Regions_df)

stargazer(Regions_model_add_abs, type = "text")

plot_model(Regions_model_add_abs, 
           type = "pred", 
           terms = c("Institutional_distrust", "age_class"),
) +
  theme_minimal()


# Interaction model of populist vote with control

Regions_model_int <- glm(populist_vote
                         ~ Macro_region + Institutional_distrust
                         * age_class,
                         family = binomial(link = "logit"),
                         data = Regions_df)

stargazer(Regions_model_int, type = "text")

plot_model(Regions_model_int, 
           type = "pred", 
           terms = c("Institutional_distrust", "age_class"),
           ci.lvl = NA
) +
  theme_minimal()

# Interaction model of abstention with control 

Regions_model_int_abs <- glm(abstention
                             ~ Macro_region + Institutional_distrust
                             * age_class,
                             family = binomial(link = "logit"),
                             data = Regions_df)

stargazer(Regions_model_int_abs, type = "text")

plot_model(Regions_model_int_abs, 
           type = "pred", 
           terms = c("Institutional_distrust", "age_class"),
) +
  theme_minimal()


# Models review (No Complete pooling)

library(modelsummary)
library(pandoc)


Populist_vote <- modelsummary(
  list(
    "Interaction model" = distrust_x_class,
    "High distrust excluded (additive)" = CDT,
    "High distrust excluded (Interaction)" = CDT_distrust_int,
    "Region controlling (Additive)" = Regions_model_add,
    "Region controlling (Interaction)" = Regions_model_int
  ),
  output = "Populist_regression.docx",
  stars = TRUE,
  gof_omit = "IC|Log|Adj|Within|Pseudo",
)


abstention <- modelsummary(
  list(
    "Interaction model" = distrust_x_class_abs,
    "Region controlling (Additive)" = Regions_model_add_abs,
    "Region controlling (Interaction)" = Regions_model_int_abs
  ),
  output = "abstention_regression.docx",
  stars = TRUE,
  gof_omit = "IC|Log|Adj|Within|Pseudo"
)
```

## Research question and hypothesis

#### *Italian populist parties rise and fall*

In the 2018 Italian political elections 5 star movement and League for Salvini premier were the biggest winners obtaining respectively 32% and 17% of the votes; these two can be considered the main populist Italian parties . The dissatisfaction toward political parties and their system was one of the main driver for their rise [@chiaramonte2018] .

4 years later these 2 parties fell: 5 star movement obtained about 15,43% but the worst fall was the League one, collapsing to 8.8%.

#### *Younger generations and democracy*

Lots of studies in recent years showed a growing disaffection and distrust of younger generations toward democracy and its institutions and parties [e.g. @kwak2020]. This phenomenon frequently lead to two different attitudes that influence their behavior when they are called to the polls: democratic apathy and democratic antipathy; the former leading to abstention, the latter to vote for anti-system/populist parties [@foa2019].

#### *Research questions and hypothesis*

The purpose of the present study is to check the grade with which these two parties are still able to answer the distrustful and protesting political supply and whether it changes for age cohorts in the 2022 Italian general elections.

The main research question will be the following:

-   *How much the distrust toward political and party system translates in voting for populist parties? How does It changes for different cohorts?*

It will be articulated in 2 main hypothesis:

**H1**: Higher level of distrust toward Parliament and political parties will tend to significantly increase the share of votes for populist parties.

**H2**: The positive effect of institutional distrust on populist voting will be weaker among younger cohorts (18-34), who are more likely to convert distrust into abstention rather than partisan choice.

## Data and variable operationalization <!--  150 words -->

#### *Data source*

To operationalize distrust and its relation with voting behavior the ITANES post-electoral data from 2022 elections [@vezzoni2023] has been employed; it provides individual level observations for 4696 individuals.

#### *Selected variables and operationalization*

```{r}
#| include: FALSE
#| label: Trust variables


```

*Institutional distrust*: To operationalize this dimension an index was built. To do so factor analysis and principal component analysis were performed on 5 selected items. The analyses highlighted 2 main dimensions: institutional distrust and stealth democracy, both theoretically correlated with populism. Nevertheless only the first dimension was employed for the present study as it was better defined and, theoretically, a predictor for both populism and abstention. More on the indexes construction in the [appendix A](#sec-appendixA).

Institutional distrust is composed by three items: trust toward parliament, government and parties. This three items had range 1-4 were 1 meant a lot of trust. I then constructed a weighted index and rescaled it 0-10 for interpretability.

```{r}
#| include: false
#| label: Indexes creation

# Create a var vector for convenience

cat_var <- c("Q04b", "Q05b", "Q05c", "Q07a", "Q07b","Q07e") 

# With mixcor we impute a mixed correlation matrix

mixed_cormat <- mixedCor(data = trust_data,
                         c = cat_var,
                         smooth = TRUE)

# Plot the correlation 

corrplot(mixed_cormat$rho, method = "number",
         type = "lower", order = "hclust",
         tl.col = "black", tl.srt = 45)

# Since there are a lot of strong correlation use pca to determine how many 
# dimension to use 

eigenvalues <- eigen(mixed_cormat$rho)$values

explained_variance <- (eigenvalues / sum(eigenvalues)) * 100

df_screeplot <- data.frame(
  Component = 1:length(eigenvalues),
  Eigenvalues = eigenvalues,
  Variance <- explained_variance
)

print(df_screeplot)

# We have 3 component with >1 eigenvalue, so for the Kaiser method (keeping PCs)
# with EV > 1 we would take 3 of them. Anyway, we can also use a screeplot to visualize it

ggplot(df_screeplot, aes(x = as.factor(Component), y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue", width = 0.6) +
  geom_line(aes(x = Component, y = Variance), color = "black", linewidth = 0.8) +
  geom_point(aes(x = Component, y = Variance), color = "red", size = 3) +
  geom_hline(yintercept = (1/length(eigenvalues))*100, linetype = "dashed", color = "darkgray") +
  geom_text(aes(label = round(Variance, 1)), vjust = -1.5, size = 3.5) +
  labs(title = "Scree Plot",
       subtitle = "Elbow method",
       x = "Components",
       y = "Explained variance") +
  theme_minimal()

# The screeplot shows a sharp cut between first and second component, suggesting that
# We can try both 1 & 3 components

pca_model <- principal(r = mixed_cormat$rho,
                       nfactors = 1,
                       rotate = "varimax")

pca_model$loadings
pca_model$uniquenesses

pca_model2 <- principal(r = mixed_cormat$rho,
                        nfactors = 2,
                        rotate = "varimax")

pca_model2$loadings
pca_model2$uniquenesses
pca_model2$weights


# Perform FA with 2 dimension 
fa <- fa(r = mixed_cormat$rho,
         nfactors = 2,
         rotate = "varimax")

# Summarize fa and extract loadings & scores
summary(fa)
fa$loadings
fa$uniquenesses
fa$weights

# We now discovered 2 indexes: institutional distrust (MR1) & Stealth democracy attitudes (MR2)
# Stealth democracy: want the system to work well but don't want the politics to 
# lead it

# I WON'T INCLUDE STEALTH DEMOCRACY IN MY COMPUTATION

#Create the new weighted indexes & scale it 0-10

trust_data <- trust_data %>%
  mutate(Institutional_distrust = Q04b*-0.002 + Q05b*-0.036 + Q07a*0.501 + 
           Q05c*-0.006 + Q07b*0.331 + Q07e*0.211)

#we determine a function to rescale -5 +5

rescale <- function(x) {
  x_numeric <- as.numeric(x) #force x to be numeric
  
  x_min <- min(x_numeric, na.rm = TRUE)
  x_max <- max(x_numeric, na.rm = TRUE)
  
  (((x_numeric - x_min) * 10) / (x_max - x_min))
}

trust_rescaled <- trust_data %>%
  mutate(Institutional_distrust = rescale(Institutional_distrust),
  )%>%
  select(seriale, Institutional_distrust)
```

*Populist vote*: votes for 5SM and League were coded as populist and other as non populist. The selection of this two come from CHES & POPPA [@rovny2025; @zaslove2024] scores, two of the main expert surveys about political parties[^1] .

[^1]: In both surveys Brothers of Italy was the third in populist score; its inclusion could be more complete but also a source of noise as with 26% of the preferences Giorgia Meloni's party attracted a very wide sector of the electorate not only the disillusioned.

*Abstention*: for abstention the voted/non voted variable from ITANES was employed

*Age class*: the survey presented a 6 class age classification which were recoded to 5 for 2 reasons:

1.  The main hypothesis is about younger generations which include also 25-34. The study main model will be a logistic model, which require a reference category; having them both as reference category will help to compare them directly with other cohorts;
2.  19-24 and 24-35 are the two less numerous categories, specially the lower one which presented less than a half observation respect to the others. Merging them was useful to have stronger predictive power.

```{r}
#| include: false
#| label: final df creation

#FINAL DF

populist_df <- ITANES_2022_Post_electoral%>%
  select(seriale, eta_classe, Q10LHa, Q10LHb,eta)%>%
  mutate(seriale = as.numeric(zap_labels(seriale)),
         abstention = as.numeric(zap_labels(Q10LHa)))%>%
  filter(abstention < 3)%>%
  mutate(abstention = ifelse(abstention == 2, 1, 0),
         abstention = as.factor(abstention))%>%
  mutate(populist_vote = Q10LHb %in% c("2","5"))%>%
  mutate(eta_classe = fct_collapse( #merging classes 1 & 2
    factor(eta_classe),
    "18-34" = c("1", "2"),  
    "35-44"   = "3",
    "45-54"   = "4",
    "55-64"   = "5",
    ">64"   = "6"
  ))%>%
  mutate(age_class = eta_classe)%>%
  select(seriale, age_class, populist_vote, eta, abstention)


pop_age <- left_join(trust_rescaled, populist_df, by = join_by(seriale))

```

*Macro Region*: for electoral behavior geographical positioning is always a strong predictor in Italy, observations were divided them in the 3 classical macro regions: north, south and red zone (Toscana, Emilia-Romagna, Marche, Umbria).

```{r}
#| include: false
#| label: regions-coding

Regions_df <- ITANES_2022_Post_electoral%>%
  mutate(Macro_region = case_when(
    regione < 7 ~ "North",
    regione < 12 ~ "Red_area",
    TRUE ~ "South"))%>%
  select(seriale, Macro_region)

Regions_df <- left_join(pop_age_clean, Regions_df, by = join_by(seriale))%>%
  mutate(Macro_region = factor(Macro_region))
```

More information about employed items and their item-code are available in [appendix B](#sec-appendixB).

## Model choice and statistical justification <!-- 300 words -->

#### *Main model*

To test H1 a logistic regression model has been employed for both populist vote and abstention. This kind of regression is appropriate as it models the outcome as a probability within the \[0,1\] range and both the dependent variables are binary. Main independent variable was institutional distrust.

The initial univariate analysis has been performed and it led to a small effect, suggesting potential misspecification of the model, particularly the violation of the linearity assumption in the logit.

Since the logistic regression failed to produce strong significative results a Generalized addittive model (GAM) was employed as a diagnostic tool to check if the fitting problem was due to data distribution; it was ideal because it doesn't assume linearity and adapt to the shape of the data and reveal potential non linear patterns. This showed a non-monotonic relationship between distrust and populist voting with a collapse in populist voting for high levels of distrust, and, on the contrary, abstention displays a consistently positive relationship.

Based on this evidence a restricted specification, labelled Critical Distrust Threshold (CDT), was introduced. It restricted the sample respondents below the threshold of 8. A better fitting on this data would confirm that distrust works as a positive predictor till medium-high levels of distrust.

#### *Interaction and Additive model*

To test H2 age class was added and incorporated in both additive and interaction models. The aim of the first one was to study the effect of age class isolated; the second to check whether distrust interacted with age class, trying to have a better comprehension of how distrust translates to electoral behavior for each age class. The selected reference category has been 18-34, as it was the main focus of the study: this choice permitted to compare it to all the other classes.

Lastly a control for Macro Region was included, to account for persistent territorial differences in voting behavior.

## Results <!-- 500 words -->

```{r}
#| label: Full-Model 
#| include: false

#Addittive models-----

pop_age$populist_vote <- as.factor(pop_age$populist_vote)

Complete_pooling <- glm(populist_vote ~ Institutional_distrust,
                        family = binomial(link = "logit"),
                        data = pop_age)

stargazer(Complete_pooling, type = "text")

plot_model(Complete_pooling, 
           type = "pred", 
           terms = c("Institutional_distrust"),
) +
  theme_minimal()

#Try it on vote/non vote

CP_abstention <- glm(abstention ~ Institutional_distrust,
                     family = binomial(link = "logit"),
                     data = pop_age)

stargazer(CP_abstention, type = "text")

plot_model(CP_abstention, 
           type = "pred", 
           terms = "Institutional_distrust",
) +
  theme_minimal()

#try fitting the model only on voters (exclude abstention)

no_abs <- pop_age%>%
  mutate(abstention = as.numeric(abstention))%>%
  filter(abstention<2)

CP_voters <- glm(populist_vote ~ Institutional_distrust,
                 family = binomial(link = "logit"),
                 data = no_abs)

stargazer(CP_voters, type = "text")

plot_model(CP_voters, 
           type = "pred", 
           terms = "Institutional_distrust",
) +
  theme_minimal()



# INTERACTION MODEL ----

# Clean the dataset to perform interaction model and avoid confusing sjPlot

pop_age_clean <- pop_age %>%
  filter(Institutional_distrust > 0)%>%
  drop_na() %>%
  mutate(
    log_distrust = log1p(Institutional_distrust), #without transforming them before sjPlot wouldn't recognize them so we have to perform this now
    age_class   = as.factor(age_class)
  )

# INSTITUIONAL DISTRUST X AGE CLASS

# I have to decide a reference class. For my purpose the main strategy would be to choose 
# 18-34 to see its behaviour confronted with the others 

table(pop_age_clean$age_class)  

#18-34 35-44 45-54 55-64   >64 
#756   701   909   852    1269 

# A solution could be to merge 1 & 2

pop_age_clean$age_class <- relevel(factor(pop_age_clean$age_class), ref = "18-34")

distrust_x_class <- glm(populist_vote
                        ~ Institutional_distrust
                        * age_class,
                        family = binomial(link = "logit"),
                        data = pop_age_clean)

stargazer(distrust_x_class, type = "text")

sjPlot::plot_model(distrust_x_class, 
                   type = "int",
                   show.legend = TRUE
) +
  theme_bw()

distrust_x_class_abs <- glm(abstention
                            ~ Institutional_distrust
                            * age_class,
                            family = binomial(link = "logit"),
                            data = pop_age_clean)

stargazer(distrust_x_class_abs, type = "text")

sjPlot::plot_model(distrust_x_class_abs, 
                   type = "int",
                   show.legend = TRUE
) +
  theme_bw()

# GAM - Check the shape of the distribution ----

# A gam is a model that guarantee a high flexibility & calculate non linear regression models
# we want to use it to study the shape of the relation

library(mgcv)

mod_gam <- gam(populist_vote
               ~ s(Institutional_distrust)
               + age_class,
               family = binomial, data = pop_age_clean)

plot(mod_gam, pages = 1, shade = TRUE)

summary(mod_gam)

# It seems like at a certain point the populist vote decrease for institutional distrust
# This can suggest that after a certain ceiling non voting become the choice: this could
# be the reason for our non working models. We can try a model that take institutional
# distrust till 7/8 and check the significativity of the relation and we can do the same
# for non vote. Moreover we can impute a GAM for abstention


mod_gam <- gam(abstention
               ~ s(Institutional_distrust)
               + age_class,
               family = binomial, data = pop_age)

plot(mod_gam, pages = 1, shade = TRUE)

# Our HP was right abstention explode for really high levels of distrust 
# Now we can try our CDT (distrust but not so much) model with level lower than 8

# Before starting with our plan we want to check how many high distrust individuals we have
table(pop_age$Institutional_distrust >= 8)

# 987 is really significant, this means that they have a strong impact on our regression models

# CDT Model -----

# CDT addictive

CDT_df <- pop_age_clean%>%
  filter(Institutional_distrust <= 8)

CDT <- glm(populist_vote
             ~ Institutional_distrust,
             family = binomial(link = "logit"),
             data = CDT_df)

stargazer(CDT, type = "text")

plot_model(CDT,
           type = "pred",
           terms = "Institutional_distrust",
           show.legend = TRUE) + 
  theme_minimal()

# CDT Interaction 

CDT_df$age_class <- relevel(factor(CDT_df$age_class), ref = "18-34")

CDT_distrust_int <- glm(populist_vote
                          ~ Institutional_distrust
                          * age_class,
                          family = binomial(link = "logit"),
                          data = CDT_df)

stargazer(CDT_distrust_int, type = "text")

plot_model(CDT_distrust_int, 
           type = "int",
           ci.lvl = NA,
           show.legend = TRUE,
) +
  theme_bw()

table(pop_age_clean$age_class)     


# Control by FDI
# Control by macro-regions / instruction 
# Control by M5S/League

#Macro regions controlling ----

# North - South - Red zone (included: Toscana, Emilia-Romagna, Umbria, Marche)

Regions_df <- ITANES_2022_Post_electoral%>%
  mutate(Macro_region = case_when(
    regione < 7 ~ "North",
    regione < 12 ~ "Red_area",
    TRUE ~ "South"))%>%
  select(seriale, Macro_region)

Regions_df <- left_join(pop_age_clean, Regions_df, by = join_by(seriale))%>%
  mutate(Macro_region = factor(Macro_region))

# Selection of the reference class

Regions_df$age_class <- relevel(factor(Regions_df$age_class), ref = "18-34")

#Pure addittive model of populist vote with control

Regions_model_add <- glm(populist_vote
                         ~ Macro_region + Institutional_distrust
                         + age_class,
                         family = binomial(link = "logit"),
                         data = Regions_df)

stargazer(Regions_model_add, type = "text")

plot_model(Regions_model_add, 
           type = "pred", 
           terms = c("Institutional_distrust", "age_class"),
) +
  theme_minimal()

# Pure addittive model of abstention with control


Regions_model_add_abs <- glm(abstention
                             ~ Macro_region + Institutional_distrust
                             + age_class,
                             family = binomial(link = "logit"),
                             data = Regions_df)

stargazer(Regions_model_add_abs, type = "text")

plot_model(Regions_model_add_abs, 
           type = "pred", 
           terms = c("Institutional_distrust", "age_class"),
) +
  theme_minimal()


# Interaction model of populist vote with control

Regions_model_int <- glm(populist_vote
                         ~ Macro_region + Institutional_distrust
                         * age_class,
                         family = binomial(link = "logit"),
                         data = Regions_df)

stargazer(Regions_model_int, type = "text")

plot_model(Regions_model_int, 
           type = "pred", 
           terms = c("Institutional_distrust", "age_class"),
           ci.lvl = NA
) +
  theme_minimal()

# Interaction model of abstention with control 

Regions_model_int_abs <- glm(abstention
                             ~ Macro_region + Institutional_distrust
                             * age_class,
                             family = binomial(link = "logit"),
                             data = Regions_df)

stargazer(Regions_model_int_abs, type = "text")

plot_model(Regions_model_int_abs, 
           type = "pred", 
           terms = c("Institutional_distrust", "age_class"),
) +
  theme_minimal()


# Models review (No Complete pooling)

library(modelsummary)
library(pandoc)


Populist_vote <- modelsummary(
  list(
    "Interaction model" = distrust_x_class,
    "High distrust excluded (additive)" = CDT,
    "High distrust excluded (Interaction)" = CDT_distrust_int,
    "Region controlling (Additive)" = Regions_model_add,
    "Region controlling (Interaction)" = Regions_model_int
  ),
  output = "Populist_regression.docx",
  stars = TRUE,
  gof_omit = "IC|Log|Adj|Within|Pseudo",
)


abstention <- modelsummary(
  list(
    "Interaction model" = distrust_x_class_abs,
    "Region controlling (Additive)" = Regions_model_add_abs,
    "Region controlling (Interaction)" = Regions_model_int_abs
  ),
  output = "abstention_regression.docx",
  stars = TRUE,
  gof_omit = "IC|Log|Adj|Within|Pseudo"
)

```

#### *Institutional distrust on populist vote and abstension*

The pure

## Interpretation and substantive insights <!-- 300 words -->

## Limitations and reflection <!-- 300 words -->

#### Appendix A {#sec-appendixA .appendix}

#### Appendix B {#sec-appendixB .appendix}

#### Bibliography
